Density-dependent selection with strictly relative contests

### 2017-11-16 manuscript.tex
l

The issue can be expressed more formally by revisiting MacArthur's famous analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by general differential equations
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if either $n_1$ or $n_2$ are made sufficiently large. The outcome of selection is then determined by the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

MacArthur's argument is clearly incompatible with a narrow interpretation of relative fitness models as selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b). In this case the $f_1$ and $f_2$ nullclines are not even defined. Moreover, at face value MacArthur's conclusion that ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967} seems to be fundamentally incompatible with constant $N$. In particular, the logistic model and its bi-allelic diploid variants are widely used to analyze MacArthur's ``$K$-selection'' in which case positive selection is necessarily tied to increases in equilibrium population size. More sophisticated models using the Lotka-Volterra competition equations, or explicit resource consumption, also intrinsically tie high-density selection to changes in population density. This raises the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and experiments.

@@

Futhermore, we show that our model is entirely consistent with the claim that ``fitness is $K$'' in crowded populations provided that MacArthur's liberal use of ``$K$'' is taken into account. In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities \citep{anderson_1971}. This is largely an artifact of the models historically used in the density-dependent selection literature, which ignore relative contests.

@@

In the Introduction we described MacArthur's general conclusions about the behaviour of Eq.~\eqref{eq:macgeneral} based on the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, but did not specify the role of ``$K$''.

>>>>

The issue can be expressed more formally by revisiting MacArthur's famous analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are made sufficiently large. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the shape of these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

MacArthur's argument is often presented using the logistic model and its bi-allelic diploid variants [citations]. Positive selection in crowded populations is then necessarily tied to increases in equilibrium population size --- commonly referred to as ``$K$-selection''. Clearly this is  incompatible with the constant $N$ assumption of relative fitness models. While MacArthur's argument is not restricted to the logistic model, other standard ecological  models like the Lotka-Volterra competition equations, or consumable resource models, also tie selection in crowded populations to changes in $N$, albiet in more subtle ways (we return to this in section BLAH). More generally, it is not clear how relative fitness models fit in with MacArthur's argument at all, since they only allow for selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not even defined. We are thus faced with the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and experiments.

@@

Futhermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded poulations. In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities \citep{anderson_1971}. This is largely an artifact of the models historically used in the density-dependent selection literature, which ignore relative contests.

@@

In the Introduction we gave the essential details of MacArthur's general argument, but we did not specify the role of ``$K$''.

revolving around the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ of Eq.~\eqref{eq:macgeneral}
